---
title: "BIOS 735 Random Forest Variable Selection and Analysis"
output: html_document
author: "Ben Bodek, Brian Chen, Forrest Hurley, Brian Richardson, Emmanuel Rockwell"
date: "04/28/22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages(if (!require(dplyr)){install.packages("dplyr")})
suppressPackageStartupMessages(if (!require(tidyr)){install.packages("tidyr")})
suppressPackageStartupMessages(if (!require(ggplot2)){install.packages("ggplot2")})
suppressPackageStartupMessages(if (!require(ordinalNet)){install.packages("ordinalNet")})
suppressPackageStartupMessages(if (!require(foreign)){install.packages("foreign")})
suppressPackageStartupMessages(if (!require(devtools)){install.packages("devtools")})
suppressPackageStartupMessages(if (!require(tictoc)){install.packages("tictoc")})
suppressPackageStartupMessages(if (!require(psych)){install.packages("psych")})
suppressPackageStartupMessages(if (!require(caret)){install.packages("caret")})
suppressPackageStartupMessages(if (!require(Boruta)){install.packages("Boruta")})
suppressPackageStartupMessages(if (!require(ordinalForest)){install.packages("ordinalForest")})
suppressPackageStartupMessages(if (!require(cvms)){install.packages("cvms")})
suppressPackageStartupMessages(if (!require(ggplot2)){install.packages("ggplot2")})
suppressPackageStartupMessages(if (!require(data.table)){install.packages("data.table")})
suppressPackageStartupMessages(if (!require(rlang)){install.packages("rlang")})
load_all()

#For reproducibility
set.seed(1)
```

# Read in and Process Data

```{r}
# Note that data must be downloaded and processed (instructions in data_processing.R)
# prior to running this step

# reading in processed data
data <- fread('processed_data.csv')

y_factor_key <- data.frame(scaled = sort(unique(data$`B1:SATISFACTION W/LIFE`)),
                           y = factor(1:5, ordered = T))

data_ordered_y <- merge(y_factor_key, data, by.x = "scaled", by.y = "B1:SATISFACTION W/LIFE") %>% 
  arrange(`W1 Case Id`) %>%
  select(-c("scaled", "W1 Case Id"))
```

We start by defining our outcome and covariates. Our outcome of interest is the participants' **Satistfaction with Life**, an ordinal survey response ranging from 1 = "Completely Satisfied" to 5 = "Not at all Satisfied".

The covariates are defined as all other survey questions. 

```{r}
# set outcome as Life Satisfaction, "B1.SATISFACTION.W.LIFE"
y<-data$`B1:SATISFACTION W/LIFE`
# create covariate matrix
x<-select(data, select = -c("B1:SATISFACTION W/LIFE"))
```


Below is a table of percentage of responses by ordinal category. We can see that most responses
fall into category 2, "Very satisfied with life".

```{r}
l<-table(y)/sum(!is.na(y))*100
barplot(l)
```
```{r}
set.seed(5)

#Create training and testing set
train_test_i <- createDataPartition(y = data_ordered_y$y, p = 0.6, list = F)

data_train <- data_ordered_y[train_test_i,]
data_test <- data_ordered_y[-train_test_i,]

#Within training set, create folds
k <- 5 #set number of folds 
fold_indicies <- createFolds(y = data_train$y, k=k)
```

# Variable Selection with Random Forest on training set
```{r}

#Fit Ordinal Random forest
tic()
rf<-randomforest.features("y",data_train)
toc()
  
# extract important features
imp.features<-rf$important.features

# train model on selected features using training dataset
model<-ordinalForest::ordfor("y", data_train[,c("y",imp.features)])

```

# Evaluate model using test set
```{r}
# save predictions on test set
predicted <- predict(model,data_test[,-1])
    
#Calculate Kappa
test_kappa <- cohen.kappa(x = matrix(c(data_test[,1], predicted[-length(predicted)]),
                                           ncol = 2, 
                                           byrow = F))$weighted.kappa
# creating and plotting a confusion matrix

# save actual y and predicted y into a tibble
y = data_test[,1]
y.pred = predicted[-length(predicted)]$ypred
y.pred <- factor(y.pred, order=T)
d_multi<-tibble("target"=y,"prediction"=y.pred)

conf_mat<-confusion_matrix(targets = d_multi$target,
                             predictions = d_multi$prediction)
plot_confusion_matrix(conf_mat$`Confusion Matrix`[[1]],add_normalized=F)
```

# Plotting selected important features for intuition

```{r message=FALSE}
# function to create plots for important categorical features
plt_important_cat_features<-function(df,feature,file_location){
  # create data frame for plotting
  plt_dat<-df%>%select(c(`B1:SATISFACTION W/LIFE`,!!sym(feature)))
  # round the feature and filter for values >=1 if there are more than 2 response options
  # (some imputed features may have erronious values <1)
  plt_dat[,feature]<-round(plt_dat[,..feature])
  if(nrow(unique(data[,..feature]))>2){
    plt_dat<-plt_dat%>%filter(!!sym(feature)>=1)}
  # group data to prepare for graphing
  plt_dat<-plt_dat%>%group_by(!!sym(feature),`B1:SATISFACTION W/LIFE`)%>%
    summarise(count=n())%>%mutate(prcnt=100*count/sum(count))%>%ungroup
  # create plot object
  plt<-ggplot(plt_dat,aes(x = factor(`B1:SATISFACTION W/LIFE`),y=prcnt,fill=factor(`B1:SATISFACTION W/LIFE`)))+
    geom_bar(stat="identity")+
    ggtitle(paste0("Plot of 'SATISFACTION W/ LIFE (1=high)' \n by '",feature,"'"))+
    ylab("Percent of Responses")+
    xlab("")+
    labs(fill="'SATISFACTION W/ LIFE (1=high)'")+
    theme_bw()+
    theme(panel.grid.major.x = element_blank())+
    facet_grid(paste0(". ~ `",feature,"`"))
  # save plot as png
  ggsave(paste0(file_location,substr(sub(":.*", "",feature),1,4),"_plot.png"),plt,width=10,height=6)
}
# function to create heatmaps for important categorical features
plt_heatmap<-function(df,feature,file_location){
  # create data frame for plotting
  plt_dat<-df%>%select(c(`B1:SATISFACTION W/LIFE`,!!sym(feature)))
  # round the feature and filter for values >=1 (some imputed features may have erronious values <1)
  plt_dat[,feature]<-round(plt_dat[,..feature])
  if(nrow(unique(data[,..feature]))>2){
    plt_dat<-plt_dat%>%filter(!!sym(feature)>=1)}
  # group data to prepare for graphing
  plt_dat<-plt_dat%>%group_by(!!sym(feature),`B1:SATISFACTION W/LIFE`)%>%
    summarise(count=n())%>%ungroup
  # create plot object
  plt<-ggplot(plt_dat,aes(y = factor(`B1:SATISFACTION W/LIFE`),x=factor(!!sym(feature)),fill=count))+
    geom_tile(color="black")+
    ggtitle(paste0("Plot of 'SATISFACTION W/ LIFE (1=high)' \n by '",feature,"'"))+
    xlab(feature)+
    ylab('SATISFACTION W/ LIFE')+
    labs(fill="Count of respondents")+
    scale_fill_gradient(low="white",high="red")+
    coord_fixed()
  # save plot as png
  ggsave(paste0(file_location,substr(sub(":.*", "",feature),1,4),"_heatmap.png"),plt)
}

plt_important_cont_features<-function(df,feature,file_location){
  plt_dat<-df%>%select(c(`B1:SATISFACTION W/LIFE`,!!sym(feature)))
  plt<-ggplot(plt_dat,aes(x= factor(`B1:SATISFACTION W/LIFE`),y=!!sym(feature),fill=factor(`B1:SATISFACTION W/LIFE`)))+
    geom_boxplot()+
    ggtitle(paste0("Plot of 'SATISFACTION W/ LIFE (1=high)' \n by '",feature,"'"))+
    ylab(feature)+
    xlab("'SATISFACTION W/ LIFE (1=high)'")+
    labs(fill="'SATISFACTION W/ LIFE (1=high)'")+
    theme_bw()
  # save plot as png
  ggsave(paste0(file_location,substr(sub(":.*", "",feature),1,4),"_plot.png"),plt,width=10,height=6)
}

# create and save plots
for(f in imp.features){
  # check if variable is categorical
  if(nrow(unique(data[,..f]))<=15){
    plt_important_cat_features(data,f,"rf plots/")
    plt_heatmap(data,f,"rf plots/")}
  else{
    plt_important_cont_features(data,f,"rf plots/")
    }
}
```